Optimal forward modeling of gravitational fields using tesseroids with adaptive discretization

### 2014-08-21 manuscript.tex
Updated split fig to split in each dimension
Better figure for tessXshell
Font size is still off and need a better color palette. It looks pretty
ugly as is.
Tweaks to tessXshell figure
Added a matplotlibrc file with the nice config for the plot.
More text in the methods
Moved some things to the introduction. Stopped before the division
algorithm  description.

@@

\noindent
in which
$(\phi, \lambda, r)$ are
the latitude, longitude, and radius
coordinates of computation point P (Fig.~\ref{fig:tesseroid}),
$\alpha,\beta \in \{x,y,z\}$
(the local coordinate system of point P),
and

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figs/tesseroid}
    \caption{
        View of a tesseroid,
        the integration point $Q$,
        a geocentric coordinate system $(X, Y, Z)$,
        the computation $P$ and it's local coordinate system $(x, y, z)$.
        $r$, $\phi$, $\lambda$ are
        the radius, latitude, and longitude, respectively, of point $P$,
        and $\ell$ is the Cartesian distance between $P$ and $Q$.
    }
    \label{fig:tesseroid}
\end{figure}

The above integrals
have to be solved numerically
\citep{Wild-Pfeiffer2008}
and different approaches have been proposed.
\citet{Heck2007}
used a Taylor series expansion,
while \citet{Asgharzadeh2007}
used the Gauss-Legendre Quadrature (GLQ).
\citet{Wild-Pfeiffer2008} investigated
the use of both methods,
along with alternative mass elements,
and found the GLQ to be the most efficient method.
Therefore, we have opted to use
the 3D GLQ in our computations.

The Gauss-Legendre Quadrature
approximates the integral by
a weighted sum
\citep{Hildebrand1987},
e.g.

@@

\noindent
where the discretization points (nodes) $x_i$
are the roots of the $N$th order Legendre polynomial $P_N$
and the weights $W_i$ are \citep{Hildebrand1987}

\begin{equation}
    W_i = \frac{2}{(1 - x_i^2)(P'_N(x_i))^2}.
\end{equation}

@@

The accuracy of the numerical integration
using the GLQ
was investigated by \citet{Ku1977}
for the case of the right-rectangular prism.
\citet{Ku1977} found that the accuracy
depends on the ratio between
the average distance between GLQ nodes
and the distance to the observation point P.
As a rule-of-thumb,
\citet{Ku1977} suggests that
the distance between nodes
should not be greater than
the distance to point P.
\citet{Li2011} devised a scheme that
recursively subdivides the tesseroids
into smaller ones,
effectively decreasing the distance between nodes.
This scheme guarantees that
the rule-of-thumb is not violated.
Hence,
optimal accuracy of the numerical integration
is achieved automatically.

We use a modified version
of the recursive scheme of \citet{Li2011}.
We fix the number of nodes in the GLQ
$N^\lambda=N^\phi=N^r=2$.
In this case, the maximum distance between nodes
is approximately the largest dimension of the tesseroid ($L$)
\citep{Wild-Pfeiffer2008}.
For a computation point $P$
at a distance $d$ from
the center of the top face of the tesseroid
(Fig.~\ref{fig:ratio}a),
the tesseroid is divided in eight parts if

\begin{equation}
    \frac{d}{L} < R,
    \label{eq:ratio}
\end{equation}

\noindent
where $R$ is the minimum distance/size ratio desired.
In case the tesseroid is divided,
the inequality in equation \ref{eq:ratio}
is checked for each of the eight parts.
This process is repeated recursively
until no divisions are required,
ensuring that equation \ref{eq:ratio}
holds for all tesseroids created.
The gravitational effect
of the original tesseroid
is then calculated as
the sum of the effect
of the smaller tesseroids.

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using $R$ equal to
        (b) 0.5, (c) 1, and (d) 2.
        $d$ is the distance between
        the center of the top face of the tesseroid
        and $P$.
        $L$ is the largest dimension of the tesseroid.
    }
    \label{fig:ratio}
\end{figure}

@@

The methods described above
were implemented the software package
Tesseroids version 1.1.1.
The package consists
of a range of command-line programs
written in the C programming language.
Tesseroids is free and open-source software
under the terms of
the 3-clause BSD license.

Each program performs a single type of calculation.
Programs receive input from
the standard input stream (\texttt{stdin})
and command-line arguments.
Output is made in plain text
through the standard output stream (\texttt{stdout}).

The main programs in Tesseroids are
\texttt{tesspot},
\texttt{tessgx},
\texttt{tessgy},
\texttt{tessgz},
\texttt{tessgxx},
\ldots ,
\texttt{tessgzz}
(henceforth referred to collectively as \texttt{tess*} programs).
These programs
calculate the gravitational potential,
gravitational attraction,
and Marussi tensor
caused by a tesseroid model.
The computation points are passed
through the standard input stream
and the tesseroid model
is passed through
a simple model file.
All programs receive
optional command-line arguments
that control the GLQ order (default is $N^\lambda=N^\phi=N^r=2$)
and distance/size ratio $R$
for the recursive division scheme.

The complete list of programs provided in Tesseroids is:

\begin{itemize}
    \item \texttt{tesspot}, \texttt{tessgx}, \texttt{tessgy}, \texttt{tessgz},
        \texttt{tessgxx}, \texttt{tessgxy}, \texttt{tessgxz},
        \texttt{tessgyy}, \texttt{tessgyz}, \texttt{tessgzz}:
        Calculate the gravitational potential, gravitational attraction,
        and Marussi tensor caused by a tesseroid model.
    \item \texttt{prismpot},
        \texttt{prismgx}, \texttt{prismgy}, \texttt{prismgz},
        \texttt{prismgxx}, \texttt{prismgxy}, \texttt{prismgxz},
        \texttt{prismgyy}, \texttt{prismgyz}, \texttt{prismgzz}:
        Calculate the gravitational potential, gravitational attraction,
        and Marussi tensor caused by a right-rectangular prism model
        in Cartesian coordinates
        using the formula of \citet{Nagy2000}.
    \item \texttt{prismpots}, \texttt{prismgs}, \texttt{prismggts}:
        Calculate the gravitational potential, gravitational attraction,
        and Marussi tensor caused by a right-rectangular prism model
        in spherical coordinates
        using the formula of \citet{Wild-Pfeiffer2008}.
    \item \texttt{tess2prism}: Converts a tesseroid model into
        right-rectangular prisms of equal mass.
    \item \texttt{tessgrd}: Generate regular grids.
    \item \texttt{tessdefaults}: Output the default values
        and values of physical constants used.
    \item \texttt{tessmodgen}: Generate a tesseroid model
        from a gridded relief (e.g., a digital elevation model).
    \item \texttt{tesslayers}: Generate a tesseroid model of
        a series of stacked layers.
    \item \texttt{tessmass}: Calculate the total mass of a tesseroid model.
\end{itemize}

Tesseroids is freely available
online\footnote{www.leouieda.com/tesseroids}.
\textbf{(cite archive on figshare)} provides a persistent
archive for the source code and compiled binaries of Tesseroids.

@@

>>>>

The above integrals
have to be solved numerically
\citep{Wild-Pfeiffer2008}
and different approaches have been proposed in the literature.
\citet{Heck2007}
used a Taylor series expansion
of the integral kernels
around the geometrical center of the tesseroid.
\citet{Asgharzadeh2007}
used the Gauss-Legendre Quadrature (GLQ)
for the numerical integration.
\citet{Wild-Pfeiffer2008} investigated
the use of both methods,
along with alternative mass elements,
and found the GLQ to be the most efficient method.
We have opted to use
the 3D GLQ integration in our computations.

The accuracy of the numerical integration
using the GLQ
was investigated by \citet{Ku1977}
for the case of
the gravitational attraction of
a right-rectangular prism
in Cartesian coordinates.
\citet{Ku1977} states that the accuracy
depends on the ratio between
the average distance between GLQ nodes
and the distance to the observation point P.
As a rule-of-thumb,
\citet{Ku1977} suggests that
the distance between nodes
should not be greater than
the distance to point P.
Extrapolating this rule-of-thumb
to tesseroids,
\citet{Li2011}
devised a scheme that
recursively divides a tesseroid
into smaller ones when the rule is violated.
This procedure
effectively decreases the distance between nodes
and, theoretically, guarantees the accuracy of the GLQ integration.

@@

\noindent
in which
$(\phi, \lambda, r)$ are
the latitude, longitude, and radius
coordinates of computation point P (Fig.~\ref{fig:tesseroid}),
$\alpha,\beta \in \{x,y,z\}$
are the local coordinate system of point P,
and

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figs/tesseroid}
    \caption{
        View of a tesseroid,
        the integration point $Q$ inside the tesseroid,
        a geocentric coordinate system $(X, Y, Z)$,
        the computation $P$ and it's local coordinate system $(x, y, z)$.
        $r$, $\phi$, $\lambda$ are
        the radius, latitude, and longitude, respectively, of point $P$,
        and $\ell$ is the Cartesian distance between $P$ and $Q$.
    }
    \label{fig:tesseroid}
\end{figure}

We have opted to use
the 3D Gauss-Legendre Quadrature
to numerically integrate
\ref{eq:tesspot},
\ref{eq:tessgrav},
and
\ref{eq:tesstensor}.
The GLQ
approximates the integral by
a weighted sum
\citep{Hildebrand1987},
e.g.

@@

\noindent
where
$W_i$ are weights and
the discretization points (nodes) $x_i$
are the roots of the $N$th order Legendre polynomial $P_N$
scaled to the integration limits.
The roots of a second-order Legendre polynomial are
$x_1=-0.577350269$ and $x_2=0.577350269$.
For an arbitrary order $N$,
we have used the root-finder algorithm
of \citet{Barrera-Figueroa2006}
to calculate the roots of $P_N$.
The weights $W_i$ are \citep{Hildebrand1987}

\begin{equation}
    W_i = \frac{2}{(1 - x_i^2)(P'_N(x_i))^2},
\end{equation}

\noindent
where $P'_N$ is the first derivative of $P_N$.

@@

We have developed a modified version
of the recursive scheme of \citet{Li2011}.
We fix the number of nodes in the GLQ
$N^\lambda=N^\phi=N^r=2$
so that the distance between nodes
in the radial, longitudinal, and latitudinal directions
are approximately the dimensions of the tesseroid.
We define the distance $d$ as
the Cartesian distance between
the computation point $P$
and the center of the top face of the tesseroid
(Fig.~\ref{fig:ratio}a).
Thus, the rule-of-thumb of \citet{Ku1977}
can be written as

\begin{equation}
    \frac{d}{L} < r,
    \label{eq:ratio}
\end{equation}

\noindent
where
$L$ is one of the dimensions of the tesseroid
and
$r$ is the distance/size ratio.
For \citet{Ku1977} this ratio should be equal to one.
The dimensions of the tesseroid are

\begin{eqnarray*}
    L_r &=& r_2 - r_1,\\
    L_\phi &=& R(\phi_2 - \phi_1),\\
    L_\lambda &=& R(\lambda_2 - \lambda_1),
\end{eqnarray*}

\noindent
where $R$ is the mean Earth radius
and the longitudes and latitudes are given in radians.

In case the tesseroid is divided,
the inequality in equation \ref{eq:ratio}
is checked for each of the eight parts.
This process is repeated recursively
until no divisions are required,
ensuring that equation \ref{eq:ratio}
holds for all tesseroids created.
The gravitational effect
of the original tesseroid
is then calculated as
the sum of the effect
of the smaller tesseroids.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using $R$ equal to
        (b) 1, (c) 2, and (d) 6.
        $d$ is the distance between
        the center of the top face of the tesseroid
        and $P$.
        $L$ is the largest dimension of the tesseroid.
    }
    \label{fig:ratio}
\end{figure}

@@

@@

The methods described above
were implemented the software package
Tesseroids version 1.1.1.
The package consists
of a range of command-line programs
written in the C programming language.
It is freely available
online\footnote{www.leouieda.com/tesseroids}
under the terms of
the open-source BSD 3-clause license.
\textbf{(cite archive on figshare)} provides a persistent
archive for the source code and compiled binaries of Tesseroids.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figs/tesseroid-x-shell}
    \caption{
        Tesseroid vs spherical shell.
    }
    \label{fig:tesseroid-x-shell}
\end{figure*}
